---
title: "Analysis of Ontario wages in relation to economic factors based on Ontario Data Catalogue (1997-2019)"
author:
  - "Borys Łangowicz (1010725967)"
  - "Kian Dianati (1010205485)"
date: "2024-04-05"
output:
  pdf_document:
    number_sections: true
    latex_engine: xelatex
  html_document:
    number_sections: true
---

```{r, include=F}
library(tidyverse)
library(ggplot2)
library(reshape2)
library(knitr)
library(randomForest)
library(dplyr, warn.conflicts = FALSE)
options(dplyr.summarise.inform = FALSE)
```

# Loading the Data

We will use the following data sets:

```{r, warning=FALSE}
wages = read.csv("wages.csv") %>% mutate_if(is.character, str_trim)

wages$Education.level =
  factor(wages$Education.level,
         levels = c("Above bachelor's degree",
                    "Bachelor's degree",
                    "University certificate below bachelors degree",
                    "University degree",
                    "Community college, CEGEP",
                    "Trade certificate or diploma",
                    "Post-secondary certificate or diploma",
                    "Some post-secondary",
                    "High school graduate",
                    "Some high school",
                    "PSE  (5,6,7,8,9))",
                    "No PSE  (0,1,2,3,4)",
                    "0 - 8  years",
                    "Total, all education levels"),
         ordered = TRUE)

wages$Age.group =
  factor(wages$Age.group,
         levels = c("25-64 years",
                    "25-54 years",
                    "25-34 years",
                    "20-34 years",
                    "15-24 years",
                    "55 years and over",
                    "25 years and over",
                    "15 years and over"),
         ordered = TRUE)

# Fuel
fuel = read.csv("fuel.csv") %>% mutate_if(is.character, str_trim)

fuel <- fuel %>% rename(
  Toronto.West = Toronto.West.Ouest,
  Toronto.East = Toronto.East.Est,
  St.Catharine = St..Catharine.s,
  Ontario.Average = Ontario.Average.Moyenne.provinciale,
  Southern.Average.Ontario = Southern.Average.Moyenne.du.sud.de.l.Ontario,
  Northern.Average.Ontario = Northern.Average.Moyenne.du.nord.de.l.Ontario 
) %>% select(!(Type.de.carburant)) 
```

# Description of the Data set

## Wages by education level 
The `wages` data set includes the average weekly wages rates by education level 
and immigration status for Canada and Ontario in the years from 1997 to 2019. 
It includes the following columns:

```{r, warning=FALSE, echo=FALSE}
names(wages)
```

1. `YEAR`: Indicates the year in which the data was collected.
2. `Geography`: Indicates the region from which the data was collected. Its 
   possible values include Canada as well as the Canadian provinces and 
   territories.
3. `Type.of.work`: Indicates whether the data in the row is for full-time 
   employees or part-time employees or both.
4. `Wages`:
    1. `Total employees`: The number of employees in the given age range, 
       education level, and job status.
    2. `Average hourly wage rate`: The average hourly wage of the employees in 
       the given age range, education level, and job status.
    3. And so on for `Average weekly wage rate`, `Median hourly wage rate`, 
       and `Median weekly wage rate`.
5. `Education.level`: Indicates the level of education. It can include the 
   following:
```{r, echo=FALSE, warning=FALSE}
kable(wages %>%
  group_by(Education.level) %>%
  reframe(Education.level) %>%
  unique())
```
6. `Age.group`: Indicates the age range of the individuals under consideration.
   It can include the following:
```{r, echo=FALSE, warning=FALSE}
kable(wages %>%
  group_by(Age.group) %>%
  reframe(Age.group) %>%
  unique())
```
7. `Both.sexes`: The data not seperated by gender.
8. `Male`: The data for males.
9. `Female`: The data for females.

## Fuels price survey information 
```{r, warning=FALSE, echo=FALSE}
names(fuel)
```

1. `Date`: Indicates the date on which the data was collected.
2. `Fuel Price`: Represents the price of fuel.
3. `Ottawa`, `Toronto.West`, `Toronto.East`, `Windsor`, `London`, `Peterborough`, `St.Catharine`, `Sudbury`, `Sault.Saint.Marie`, `Thunder.Bay`, `North.Bay`, `Timmins`, `Kenora`, `Parry.Sound`: Represents the fuel price in various locations in Ontario, Canada.
4. `Ontario.Average`: Indicates the average fuel price across different regions of Ontario.
5. `Southern.Average.Ontario`: Indicates the average fuel price across the southern regions of Ontario.
6. `Northern.Average.Ontario`: Indicates the average fuel price across the northern regions of Ontario.
7. `Fuel.Type`: Indicates the type of fuel associated with the data.


# The Background of the Data

The labor and demographic dataset from the Ministry of Labour, Immigration, 
Training, and Skills Development provides insights into Ontario's workforce 
demographics, including age groups, employment types, educational levels, wages,
and immigration statuses. It is annually updated and used by policymakers, 
researchers, and economists to inform decisions regarding education, training, 
workforce development, and immigration policies in the province.

Additionally, fuel price survey information from the Ministry of Energy offers 
weekly retail prices for gasoline, diesel, auto propane, and compressed natural 
gas across ten Ontario markets. This data aids in monitoring fuel price 
fluctuations and analyzing trends in the energy sector, supporting research and
analysis efforts in economics, environmental studies, and energy policy.

# Overall Research Question

Our research aims to explore wage dynamics in Ontario, considering economic factors. We're particularly interested in:

1. How has the average hourly wage rate changed over the years across different age groups?
2. How does the average hourly wage rate differ across various education levels?
3. Are there any trends in the employment rates based on different levels of education attainment?
4. How do wage rates vary across different age groups, and is there a trend in wage growth as individuals age?
5. Is there a significant gender wage gap, and how has it evolved over time?
6. How does educational attainment affect the gender wage gap?
7. How has the total number of employees changed over the years?
  
# Analysis

## How has the average hourly wage rate changed over the years across different age groups?

```{R, echo=F}
avg_wage_by_age <- wages %>%
  filter(Wages == "Average hourly wage rate",
         Geography == "Canada",
         Type.of.work == "Full-time",
         Education.level == "Total, all education levels") %>%
  select(YEAR, Age.group, Both.Sexes) %>%
  group_by(YEAR, Age.group) %>%
  summarise(Avg_Hourly_Wage = mean(Both.Sexes))

kable(avg_wage_by_age %>%
  pivot_wider(names_from = Age.group, values_from = Avg_Hourly_Wage))
```

This table shows the average wage of each age group in dollars per hours for
the given years. For instance, in the year 1997 those who were between 15 and 24
years of age made 9 dollars and 82 cents per hour of work while those who were
25 to 34 years old made 15 dollars and 29 cents.

This table indicates that there has been a slow but steady growth in wages
accross all age groups over the years from 1997 to 2019. Moreover, it appears
that older age groups consistently earn higher wages than those in younger ones.
For instance, in 2019 the average wage for those who were older than 55 was 30 
dollars and 12 cents while those who were between 15 and 24 made only 19 dollars
and 32 cents. However, the average wage of the oldest age group, 55+, was not
much higher than the average wage of the total population—which in 2019 was 
29.56 dollars per hour. This suggests that the number of employees between 15
and 24 years of age is insignificant relative to older age groups.

**These results can also be seen in the following bar chart**:

```{r, echo=F}
ggplot(avg_wage_by_age %>% 
         filter(YEAR == "2019"),
       aes(x = Age.group, Avg_Hourly_Wage)) +
  geom_bar(stat = "identity") +
  labs(x = "Age group",
       y = "Average Hourly Wage",
       title = "Wage by Age Group in 2019") +
  theme(text = element_text(size = 8))
```

## How does the average hourly wage rate differ across various education levels for different genders?

```{R, echo=F}
avg_wage_by_education <- wages %>%
  filter(Wages == "Average hourly wage rate", 
         Geography == "Canada",
         Age.group == "15 years and over",
         # YEAR == '2019',
         Type.of.work == "Both full- and part-time") %>%
  select(Education.level, Male, Female, Both.Sexes) %>%
  group_by(Education.level) %>%
  summarise(male_avg = mean(Male),
            female_avg = mean(Female),
            Total = mean(Both.Sexes))

kable(avg_wage_by_education %>%
  mutate(pay.gap = male_avg - female_avg),
  col.names = c("Education Level", "Male", "Female", "Total", "Pay gap"),
  caption = "Average wage by education level in male and female employees")
```

**Conlusions**:
This table indicates the average hourly wage of employees based on their
education level. It also indicates the pay based on the gender of the employee.
For instance, a male with a bachelor's degree makes, on average, 28 dollars
an hour while a female makes 24 dollars resulting in a 4 dollar pay gap.

From this table and the following graph, it is clear that with higher
educational attainment the average wage rises. For example, an individual
with a high school diploma makes about 14 dollars an hour while someone
with a bachelor's degree makes 26 dollars. 

```{R, echo = F}
ggplot(avg_wage_by_education %>% melt(),
       aes(x = Education.level, fill = variable)) +
  geom_bar(aes(y = value), 
           stat = "identity", 
           alpha = 0.8,
           show.legend = TRUE,
           position = position_identity()) +
  labs(title = "Average Hourly Wage Rate by Education Level and Gender",
       x = "Education Level",
       y = "Average Hourly Wage Rate",
       fill = "Gender") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 8),
        legend.position = "bottom") +
  coord_flip()
```

**Conlusions**:

A gender pay gap is observed, with males making about 3 dollars more on average
than their female counterparts. An surprising result is that the gender pay gap 
doesn't seem to changewith education level. In the previous example, males with 
a high school degree made 3.99 dollars more than their female peers, while males
with bachelor's degrees 4.06 dollars—only a 7 cent increase. This may also be 
observed in the following plot.

```{r, echo = F}
ggplot(avg_wage_by_education %>% mutate(pay.gap = male_avg - female_avg), 
       aes(x = Education.level, y = pay.gap)) +
  geom_bar(stat = "identity") +
  labs(y = "Gender Pay Gap",
       x = "Education Level",
       title = "Gender Pay Gap by Education") +
  coord_flip()
```

**Conlusions**:

Interestingly, the largest gender gap was observed in those with a trade 
certificate, where the pay gap is almost twice as large as in other education
levels. 

These observations highlight disparities in wages between genders across 
different education levels, indicating the presence of gender-based wage 
inequality.

## What is the overall trend in the number of full-time employees versus part-time employees across different age groups?

```{R, warning=FALSE, echo = F}
full_time_part_time <- wages %>% 
  filter(Age.group %in% c("15-24 years", 
                          "25-54 years", 
                          "55 years and over"),
         Geography == 'Canada',
         Education.level == 'Total, all education levels',
         Wages == "Total employees") %>%
  group_by(Age.group, Type.of.work) %>%
  summarise(Employees = sum(Both.Sexes))

options(scipen = 999)
ggplot(full_time_part_time, 
       aes(x = Age.group, y = Employees, fill = Type.of.work)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Full-time, Part-time vs Part- and Full-time Employees\n Across Age Groups",
       x = "Age Group",
       y = "Number of Employees",
       fill = "Type of Work") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

**Conlusions**:

From this plot, we can see that the number of full-time employees is
consistently larger than the number of part-time employees across all age 
groups. In fact, this plot confirms our hypothesis from before, as we can see
that the number of employed 15-24 year olds is much smaller than the number
of employed 25-54 year-olds—who have the highest employment out of any age
group. This data suggests that as individuals grow older, they tend to favor
part-time employment. Further analysis, such as examining trends over time or 
considering demographic factors, may provide additional insights into the 
employment dynamics across different age groups.

## How has the total number of employees changed over the years?

```{r, echo = F}
employee.year = wages %>%
  filter(Wages == "Total employees",
         Geography == "Canada",
         Type.of.work == "Full-time",
         Education.level == "Total, all education levels",
         Age.group == '15 years and over') %>%
  select(YEAR, Both.Sexes)

ggplot(employee.year, aes(x = YEAR, y = Both.Sexes)) +
  geom_point() +
  labs(title = "Number of employees in a given year",
       x = "Year",
       y = "Number of Employees (thousands)")
```
This plot displays the total number of employees in a given year. For instance,
in the year 2004 there were about 11 million employees in Canada. An interesting
feature of this graph is how it captures the 2008 financial crisis, where almost
1 million Canadianas lost their jobs. What's even more interesting is that no
decrease in wages was observed amongst those who kept their jobs, as was shown
previously.

# Hypothesis Testing

**Null Hypothesis (H0)**: There is no statistically significant distinction in the average wages between male and female workers (μ_male = μ_female).

**Alternative Hypothesis (H1)**: There exists a statistically significant disparity in average wages between male and female workers (μ_male ≠ μ_female).
To examine this hypothesis, we can employ a two-sample t-test to compare the wage distributions of male and female workers. This entails segregating the dataset into two distinct groups based on gender: male and female.

The resultant p-value derived from the selected statistical test indicates the probability of observing a wage difference as extreme as, or more extreme than, the observed difference, assuming the null hypothesis holds. If the obtained p-value falls below the predetermined significance level (typically 0.05), we reject the null hypothesis, indicating a significant discrepancy in wages between male and female workers.

Moreover, by computing a confidence interval for the disparity in mean wages, we can gauge the plausible range of values for the actual difference between male and female wages.

```{R, warning=FALSE}
male_wages <- wages %>% filter(Wages == "Average weekly wage rate") %>% select(Male)
female_wages <- wages %>% filter(Wages == "Average weekly wage rate") %>% select(Female)

t.test(male_wages, female_wages)
```

**Conclusions**
The outcomes derived from the Welch Two Sample t-test underscore a significant contrast in average earnings between male and female workers. With an extraordinarily minute p-value (< 0.00000000000000022), there's compelling evidence to dismiss the null hypothesis, indicating that the mean wages for men and women diverge significantly. The 95% confidence interval for the mean wage discrepancy extends from 116.8124 to 123.6314, suggesting that the actual difference in wages between male and female workers likely lies within this interval.

In summary, the findings overwhelmingly advocate for rejecting the null hypothesis, underscoring a noteworthy disparity in wages between male and female employees. Specifically, the average wage for male workers (627.8701) markedly exceeds that of female workers (507.6481).

# Bootstrapping

We utilize bootstrapping to estimate the sampling distribution of the mean fuel price for Toronto West and to compute a confidence interval.

```{R, warning=FALSE}
library(boot)

mean_fuel_price <- function(data) {
  mean(data[["Toronto.West"]])
}

set.seed(123)
n_boot <- 1000 
bootstrap_means <- replicate(n_boot, {
  sample_data <- fuel[sample(1:nrow(fuel), replace = TRUE), ]
  mean_fuel_price(sample_data)
})

ci <- quantile(bootstrap_means, c(0.025, 0.975))

# Print the results
cat("Mean fuel price for Ontario:", mean(fuel$Toronto.West), "\n")
cat("95% confidence interval:", "( ", ci[1], "-", ci[2], " )", "\n")

```

According to our findings, the mean fuel price for Ontario stands at `r mean(fuel$Toronto.West)`. Our computed 95% confidence interval spans from `r ci[1]` to `r ci[2]``. This suggests with 95% confidence that the genuine mean fuel price for Ontario lies within this interval.

# Random Forest

```{r, warning=FALSE}
d = wages %>%
  select(!Both.Sexes) %>%
  filter(Wages == 'Average hourly wage rate',
         Type.of.work == 'Both full- and part-time',
         Geography == 'Canada',
         Education.level != 'Total, all education levels',
         Age.group != '15 years and over') %>%
  pivot_longer(c(Male, Female), 
               names_to = "gender", 
               values_to = "wage") 

d = d %>% 
  mutate(group = sample(c('train', 'test'), 
                        size = nrow(d), 
                        prob = c(0.9, 0.1), 
                        replace=TRUE))

train_data = d %>% filter(group == 'train')
test_data = d %>% filter(group == 'test')

model = randomForest(wage ~ Education.level + gender + Age.group, 
                     data = train_data,
                     ntree = 10,
                     importance = TRUE)
```

# Cross Validation

```{r, warning=FALSE}
test_data$predicted = predict(model, newdata = test_data)

mean((test_data$wage - test_data$predicted)^2)
```

The MSE of around `r mean((test_data$wage - test_data$predicted)^2)` shows the average difference between actual and predicted wages. Such a value is considered significant, suggesting that the model may not generalize well to new data. This also means, that 

# Summary of Research

# Appendix
