---
title: "Analysis of Ontario wages in relation to economic factors based on Ontario Data Catalogue (1997-2019)"
author:
  - "Borys Åangowicz (1010725967)"
  - "Kian Dianati (1010205485)"
date: "2024-04-05"
output:
  html_document:
    number_sections: true
  pdf_document:
    number_sections: true
---

```{r, include=F}
library(tidyverse)
library(ggplot2)
library(knitr)
```

# Loading the Data

We will use the following data sets:

```{r}
wages = read.csv("wages.csv")
```

# Description of the Data set

The `wages` data set includes the average weekly wages rates by education level and immigration status for Canada and Ontario in the years from 1997 to 2019. It includes the following columns:

```{r}
names(wages)
```

1. `YEAR`: Indicates the year in which the data was collected.
2. `Geography`: Indicates the region from which the data was collected. Its 
   possible values include Canada as well as the Canadian provinces and 
   territories.
3. `Type.of.work`: Indicates whether the data in the row is for full-time 
   employees or part-time employees or both.
4. `Wages`:
    1. `Total employees`: The number of employees in the given age range, 
       education level, and job status.
    2. `Averag hourly wage rate`: The average hourly wage of the employees in 
       the given age range, education level, and job status.
    3. And so on for `Average weekly wage rate`, `Median hourly wage rate`, 
       and `Median weekly wage rate`.
5. `Education.level`: Indicates the level of education. It can include the 
   following:
```{r, echo=FALSE}
kable(wages %>%
  group_by(Education.level) %>%
  reframe(Education.level) %>%
  unique())
```
6. `Age.group`: Indicates the age range of the individuals under consideration.
   It can include the following:
```{r, echo=FALSE}
kable(wages %>%
  group_by(Age.group) %>%
  reframe(Age.group) %>%
  unique())
```
7. `Both.sexes`: The data not seperated by gender.
8. `Male`: The data for males.
9. `Female`: The data for females.

# Checking Data Integrity
```{R}
# Dropping rows with NA values
wages <- na.omit(wages)
# By comparison, it seems there is no NA values
```

# The Background of the Data

# Research Questions

# Tables Summary

# Graphs Summary

# Hypothesis Testing

# Bootstrapping

# Non-linear Regression Analysis

# Cross Validation

# Random Forest

# Random Forest Validation

# Summary of Research

# Appendix
